Sparse spectral and $p$-finite element methods for partial differential equations on disk slices and trapeziums

### 2019-06-11 papers/halfdisk/halfdisk.tex
New figures

\begin{figure}[t]
	\begin{subfigure}{0.32\textwidth}
	\includegraphics[scale=0.35]{sparsityoflaplacian-w11}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
	\includegraphics[scale=0.35]{sparsityofhelmholtz}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
	\includegraphics[scale=0.35]{sparsityofbiharmonic}
	\end{subfigure}
    	\caption{"Spy" plots of (differential) operator matrices, showing their sparsity. Left: the Laplace operator $\laplacewii$. Centre: the weighted variable coefficient Helmholtz operator $\laplacewii + k^2 \: T^{(0,0)\to(1,1)} \: V({J_x^{(0,0)}}^\top, {J_y^{(0,0)}}^\top) \: T_W^{(1,1)\to(0,0)}$ for $v(x,y) = 1 - (3(x-1)^2 + 5y^2)$ and $k = 200$. Right: the biharmonic operator $\biharmonic$.}
        \label{fig:sparsity}
        \centering
\end{figure}

@@

\begin{figure}[t]
	\begin{subfigure}{0.3\textwidth}
	\centering
	\includegraphics[scale=0.3]{solution-helmholtz-k=100}
	%\label{fig:solution-helmholtz}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
	\includegraphics[scale=0.5]{solutionblocknorms-helmholtz-k=20}
	\centering
        	%\label{fig:solutionblocknorms-helmholtz}
	\end{subfigure}
	\caption{Left: The computed solution to $\Delta u + k^2 \: v \: u = f$ with zero boundary conditions with $f(x,y) = xye^x$, $v(x,y) = 1 - (3(x-1)^2 + 5y^2)$ and $k = 100$. Right: The norms of each block of the computed solution of the Helmholtz equation with the given right hand side functions, with $k=20$ and $v(x,y) = 1 - (3(x-1)^2 + 5y^2)$.}
	\centering
	\label{fig:helmholtz}
\end{figure}

@@

We can tackle the problem as follows. Denote the coefficient vector for expansion of $u$ in the $\bigWii$ OP basis up to degree $N$ by $\mathbf{u}$, and the coefficient vector for expansion of $f$ in the $\bighdopii$ OP basis up to degree $N$ by $\mathbf{f}$. Since $f$ is known, we can obtain  the coefficients $\mathbf{f}$ using the quadrature rule above. We can obtain the matrix operator for the variable-coefficient function $v(x,y)$ by using the Clenshaw algorithm with matrix inputs as the Jacobi matrices ${J_x^{(0,0)}}^\top, {J_y^{(0,0)}}^\top$, yielding an operator matrix of the same dimension as the input Jacobi matrices a la the procedure introduced in \cite{olver2019triangle}. We can denote the resulting operator acting on coefficients in the $\bighdopoo$ space by $V({J_x^{(0,0)}}^\top, {J_y^{(0,0)}}^\top)$. In matrix-vector notation, our system hence becomes:
\begin{align*}
    (\laplacewii + k^2 T^{(0,0)\to(1,1)} \: V({J_x^{(0,0)}}^\top, {J_y^{(0,0)}}^\top) \: T_W^{(1,1)\to(0,0)}) \mathbf{u} = \mathbf{f}
\end{align*}
which can be solved to find $\mathbf{u}$. We can see the sparsity and structure of this matrix system in Figure \ref{fig:sparsity} with $v(x,y) = xy^2$ as an example. In Figure \ref{fig:helmholtz} we see the solution to the inhomogeneous variable-coefficient Helmholtz equation with zero boundary conditions given in (\ref{eqn:helmholtz}) in the half-disk $\Omega$, with $k=100$, $v(x,y) = (1 - (3(x-1)^2 + 5y^2))$ and $f(x,y) = xy e^x$. In Figure \ref{fig:helmholtz} we also show the norms of each block of calculated coefficients of the approximation for four right-hand sides of the inhomogeneous variable-coefficient Helmholtz equation with $k=20$ and $v(x,y) = (1 - (3(x-1)^2 + 5y^2))$ using N = 200, that is, 20,301 unknowns. The rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution. We see that we achieve algebraic convergence for the first three examples, noting that for right hand sides that vanish at the corners of our half-disk ($x=0, \: y = \pm 1$) we see faster convergence. For the final Gaussian bump example, we see that we achieve spectral convergence.

>>>>

\begin{figure}[t]
	\begin{subfigure}{0.32\textwidth}
	\includegraphics[scale=0.35]{sparsityoflaplacian-w11}
        \centering
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
	\includegraphics[scale=0.35]{sparsityofhelmholtz}
        \centering
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
	\includegraphics[scale=0.35]{sparsityofbiharmonic}
        \centering
	\end{subfigure}
    	\caption{"Spy" plots of (differential) operator matrices, showing their sparsity. Left: the Laplace operator $\laplacewii$. Centre: the weighted variable coefficient Helmholtz operator $\laplacewii + k^2 \: T^{(0,0)\to(1,1)} \: V({J_x^{(0,0)}}^\top, {J_y^{(0,0)}}^\top) \: T_W^{(1,1)\to(0,0)}$ for $v(x,y) = 1 - (3(x-1)^2 + 5y^2)$ and $k = 200$. Right: the biharmonic operator $\biharmonic$.}
        \label{fig:sparsity}
        \centering
\end{figure}

@@

\begin{figure}[t]
	\begin{subfigure}{0.3\textwidth}
	\centering
	\includegraphics[scale=0.3]{solution-helmholtz-k=100-n=1500}
	%\label{fig:solution-helmholtz}
	\end{subfigure}
	\begin{subfigure}{0.5\textwidth}
	\includegraphics[scale=0.5]{solutionblocknorms-helmholtz-k=20}
	\centering
        	%\label{fig:solutionblocknorms-helmholtz}
	\end{subfigure}
	\caption{Left: The computed solution to $\Delta u + k^2 \: v \: u = f$ with zero boundary conditions with $f(x,y) = x(1-x^2-y^2)e^x$, $v(x,y) = 1 - (3(x-1)^2 + 5y^2)$ and $k = 100$. Right: The norms of each block of the computed solution of the Helmholtz equation with the given right hand side functions, with $k=20$ and $v(x,y) = 1 - (3(x-1)^2 + 5y^2)$.}
	\centering
	\label{fig:helmholtz}
\end{figure}

@@

We can tackle the problem as follows. Denote the coefficient vector for expansion of $u$ in the $\bigWii$ OP basis up to degree $N$ by $\mathbf{u}$, and the coefficient vector for expansion of $f$ in the $\bighdopii$ OP basis up to degree $N$ by $\mathbf{f}$. Since $f$ is known, we can obtain  the coefficients $\mathbf{f}$ using the quadrature rule above. We can obtain the matrix operator for the variable-coefficient function $v(x,y)$ by using the Clenshaw algorithm with matrix inputs as the Jacobi matrices ${J_x^{(0,0)}}^\top, {J_y^{(0,0)}}^\top$, yielding an operator matrix of the same dimension as the input Jacobi matrices a la the procedure introduced in \cite{olver2019triangle}. We can denote the resulting operator acting on coefficients in the $\bighdopoo$ space by $V({J_x^{(0,0)}}^\top, {J_y^{(0,0)}}^\top)$. In matrix-vector notation, our system hence becomes:
\begin{align*}
    (\laplacewii + k^2 T^{(0,0)\to(1,1)} \: V({J_x^{(0,0)}}^\top, {J_y^{(0,0)}}^\top) \: T_W^{(1,1)\to(0,0)}) \mathbf{u} = \mathbf{f}
\end{align*}
which can be solved to find $\mathbf{u}$. We can see the sparsity and structure of this matrix system in Figure \ref{fig:sparsity} with $v(x,y) = xy^2$ as an example. In Figure \ref{fig:helmholtz} we see the solution to the inhomogeneous variable-coefficient Helmholtz equation with zero boundary conditions given in (\ref{eqn:helmholtz}) in the half-disk $\Omega$, with $k=100$, $v(x,y) = (1 - (3(x-1)^2 + 5y^2))$ and $f(x,y) = x(1-x^2-y^2)e^x$. In Figure \ref{fig:helmholtz} we also show the norms of each block of calculated coefficients of the approximation for four right-hand sides of the inhomogeneous variable-coefficient Helmholtz equation with $k=20$ and $v(x,y) = (1 - (3(x-1)^2 + 5y^2))$ using N = 200, that is, 20,301 unknowns. The rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution. We see that we achieve algebraic convergence for the first three examples, noting that for right hand sides that vanish at the corners of our half-disk ($x=0, \: y = \pm 1$) we see faster convergence. For the final Gaussian bump example, we see that we achieve spectral convergence.
